A Java Framework For Embarrassingly Parallel Problems

### 2013-11-30 paper/paper.tex
Paper mods and command line argument change
Spell checked

\begin{abstract}
An embarrassingly parallel problem, also called pleasingly parallel, is one that
can be easily broken up into identical components that do not need to interact
to find the solution. Many problems in the biological sciences are 
embarrassingly parallel but ofter their authors do not have the expertise to
implement a parallel solution. We have written a framework, implemented in Java,
that will execute the serial program in several threads. These threads working
together help cut down the total runtime of the program significantly. The next
step is to separate the framework into a front-end that will run locally and a
back-end which receives blocks of work and does the computations. This back-end
should run in the cloud, on a machine that is dedicated to running
embarrassingly parallel tasks only.
\end{abstract}

@@

Ka/Ks Calculator is not the only embarrassingly parallel program that has the
potential to take several days to run on a data set. In fact, there are several
embarrassingly parallel programs in the biological sciences. Basic Local
Alignment Search Tool, or BLAST\fotenote{blast.ncbi.nlm.nih.gov}, is an
algorithm that compares many aspects of DNA sequences, like the nucleotides or
amino acid sequences. In addition, the user can also compare to the sequences to
a database of sequences to identity known subsequences. There are many
implementations of BLAST, some of them already have parallel implementations.
Many applications however do not have parallel processing capabilities. A class
of problems that organize large amounts of data into similar groups called
clusters. Two applications, octupus\footnote{octupus.sourceforge.com} and
USEARCH\footnote{https://www.drive5.com/usearch}, are examples of programs in
this class that do not have parallel versions. With several applications that
would benefit from parallelism in this field, it was obvious it would be
desirable to develop a framework for the parallelization task.

This framework is designed to run locally but there are other options when more
power is needed. Gunarathne et al.\cite{cloud} have investigated running much
larger embarrassingly parallel biological application using various cloud
services. These services offer on-demand computational power with flexibility.
These services have built in hooks to various distributed services, such as
storage and databases. Amazon has a set of cloud computing services called
Amazon Web Services(AWS)\footnote{https://aws.amazon.com} and Microsoft has a
similar service named Microsoft Azure\footnote{https://www.windowsazure.com}.
AWS and Azure are esentially virtual instances run on machines in the cloud that
can be spun up for extra computing power and shut down when not needed.

@@

We evaluated the convenience and performace of the framework on two different
applications: the Ka/Ks Calculator and two different sequence clustering
algorithms, USEARCH and ocutupus. In all cases, we compared the framework based
application with the original using different data partitions and different
thread counts.

@@

Place holder text

@@

TODO: First paragrpah should summarize performce: its clear that framework
provices simple and effective module for adding parallelism to a class of
file-oriented EP apps. More self praise.

@@

In the future, we would like to see this work be split into a local front-end
and a cloud back-end. The configuration set-up process will run locally. The
split and merge steps depend on the cloud service chosen, but data processing
will always occur in the cloud. Splitting the framework will give performance
improvements, because the number of compute nodes can be much larger than the
number of processors on an average machine. More customization options would be
available to the end-user, such as choice of environment. This provides ultimate
control for the group to pick the best enviroment to suit its computational and
financial needs.

>>>>

\begin{abstract}
An embarrassingly parallel problem, also called pleasingly parallel, is one that
can be easily broken up into identical components that do not need to interact
to find the solution. Many problems in the biological sciences are 
embarrassingly parallel but often their authors do not have the expertise to
implement a parallel solution. We have written a framework, implemented in Java,
that will execute the serial program in several threads. These threads working
together help cut down the total runtime of the program significantly. The next
step is to separate the framework into a front-end that will run locally and a
back-end which receives blocks of work and does the computations. This back-end
should run in the cloud, on a machine that is dedicated to running
embarrassingly parallel tasks only.
\end{abstract}

@@

Ka/Ks Calculator is not the only embarrassingly parallel program that has the
potential to take several days to run on a data set. In fact, there are several
embarrassingly parallel programs in the biological sciences. Basic Local
Alignment Search Tool, or BLAST\fotenote{blast.ncbi.nlm.nih.gov}, is an
algorithm that compares many aspects of DNA sequences, like the nucleotides or
amino acid sequences. In addition, the user can also compare to the sequences to
a database of sequences to identity known subservience's. There are many
implementations of BLAST, some of them already have parallel implementations.
Many applications however do not have parallel processing capabilities. A class
of problems that organize large amounts of data into similar groups called
clusters. Two applications, octupus\footnote{octupus.sourceforge.com} and
USEARCH\footnote{https://www.drive5.com/usearch}, are examples of programs in
this class that do not have parallel versions. With several applications that
would benefit from parallelism in this field, it was obvious it would be
desirable to develop a framework for the parallelization task.

This framework is designed to run locally but there are other options when more
power is needed. Gunarathne et al.\cite{cloud} have investigated running much
larger embarrassingly parallel biological application using various cloud
services. These services offer on-demand computational power with flexibility.
These services have built in hooks to various distributed services, such as
storage and databases. Amazon has a set of cloud computing services called
Amazon Web Services(AWS)\footnote{https://aws.amazon.com} and Microsoft has a
similar service named Microsoft Azure\footnote{https://www.windowsazure.com}.
AWS and Azure are essentially virtual instances run on machines in the cloud that
can be spun up for extra computing power and shut down when not needed.

@@

We evaluated the convenience and performance of the framework on two different
applications: the Ka/Ks Calculator and two different sequence clustering
algorithms, USEARCH and ocutupus. In all cases, we compared the framework based
application with the original using different data partitions and different
thread counts.

@@

TODO: Place holder text

@@

TODO: First paragraph should summarize performance: its clear that framework
provides simple and effective module for adding parallelism to a class of
file-oriented EP apps. More self praise.

@@

In the future, we would like to see this work be split into a local front-end
and a cloud back-end. The configuration set-up process will run locally. The
split and merge steps depend on the cloud service chosen, but data processing
will always occur in the cloud. Splitting the framework will give performance
improvements, because the number of compute nodes can be much larger than the
number of processors on an average machine. More customization options would be
available to the end-user, such as choice of environment. This provides ultimate
control for the group to pick the best environment to suit its computational and
financial needs.
